---
title: "Report Context"
title-block-banner: TRUE
fig-width: 9
fig-height: 5
warning: FALSE
error: TRUE
echo: FALSE
message: FALSE
---

```{r, Data Inputs}
#| include: FALSE
library(cpaltemplates)
library(tidycensus)
library(tidyverse)
library(rio)
library(sf)
library(googlesheets4)
library(DT)

libDB <- "C:/Users/micha/CPAL Dropbox/"

googlesheets4::gs4_auth(email = "michael@childpovertyactionlab.org")

courtobs <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/1O9P06SgvZ_nRGc64W2OsSjyX9iCS_LkcsG9087HNTuk",
                               sheet = "Form Responses 1") %>%
  janitor::clean_names(.) %>%
  rename("timeStamp"="timestamp",
         "observer"="observer",
         "observationDate"="date_of_observation",
         "jpCourt"="justice_of_the_peace_court",
         "docketStart"="docket_start_time_please_indicate_what_time_the_judge_called_the_first_case_on_the_docket_use_format_hh_mm_and_specify_am_or_pm_e_g_09_32_am",
         "docketEnd"="docket_end_time_please_indicate_the_time_the_judge_finished_hearing_the_last_case_on_the_docket_use_format_hh_mm_and_specify_am_or_pm_e_g_12_34_pm",
         "caseNumber"="case_number",
         "caseType"="is_this_a_commercial_or_residential_eviction_case",
         "plaintiffName"="plaintiff_landlord_name",
         "plaintiffDesignee"="did_the_plaintiff_landlord_or_their_designee_like_a_property_manager_appear",
         "plaintiffRep"="did_the_plaintiff_landlord_have_additional_representation_of_some_kind",
         "defName"="defendant_tenant_name",
         "defAppear"="did_the_defendant_tenant_appear",
         "defRep"="did_the_defendant_tenant_have_additional_representation_of_some_kind",
         "evicReason"="reason_for_eviction_select_multiple_if_applicable",
         "totalOwned"="total_amount_owed_please_indicate_the_total_amount_that_the_judge_or_landlord_states_is_owed_include_the_sum_of_rent_or_damages_owed_plus_court_costs_use_format_xxxx_xx_for_example_1234_50",
         "noticeVacate"="was_there_any_discussion_about_the_notice_to_vacate_ntv",
         "noticeConfirm"="if_yes_did_the_judge_confirm_the_ntv",
         "jpRuled"="the_jp_ruled_in_favor_of",
         "outcomeOther"="if_you_selected_other_please_specify_the_outcome_of_the_case_below",
         "otherNotes"="additional_case_notes"
)

#dallas county evictions dataset
evictions <- rio::import(paste0(libDB, "Data Library/Dallas County/Eviction Records/Data/Dallas County Eviction Master/EvictionRecords_Master.csv"))

#court observation dates
startDate <- min(courtobs$observationDate, na.rm = TRUE)
endDate <- max(courtobs$observationDate, na.rm = TRUE)

#total days between start and end of data
totdays <- difftime(endDate, startDate, "days") %>%
  as.numeric()

#evictions by year
# most recent year (this year)
totevic1 <- evictions %>%
  filter(filed_date >= as.Date("2022-11-01") & filed_date < as.Date("2023-04-20"))

#last year
totevic2 <- evictions %>%
  filter(filed_date >= as.Date("2021-11-01") & filed_date < as.Date("2022-04-20"))

# oldest year (two years ago)
totevic3 <- evictions %>%
  filter(filed_date >= as.Date("2020-11-01") & filed_date < as.Date("2021-04-20"))

#average evictions by day, this year
avgevic1 <- round(nrow(totevic1)/totdays, digits = 1)

#average evictions by day, last year
avgevic2 <- round(nrow(totevic2)/totdays, digits = 1)

#percent change in evictions this year
chgevic1 <- round(((nrow(totevic1)-nrow(totevic2))/nrow(totevic2))*100, digits = 1)
#percent change in evictions last year
chgevic2 <- round(((nrow(totevic1)-nrow(totevic3))/nrow(totevic3))*100, digits = 1)

#average evictions by month in the last 5 years
evicMonths <- evictions %>%
  filter(filed_date >= as.Date("2019-01-01")) %>%
  group_by(lubridate::year(filed_date), lubridate::month(filed_date)) %>%
  summarize(totevic = n()) %>%
  rename(yearFiled = 1,
         monthFiled = 2) %>%
  mutate(dateFiled = as.Date(paste0(yearFiled, "-", monthFiled, "-01"))) %>%
  ungroup(.)

#month with highest number of evictions in the last 5 years
highestMonth <- evicMonths %>%
  filter(totevic == max(totevic))

#total number of SMU students observing cases
smuStudents <- n_distinct(courtobs$observer)

#total number of cases observed
totObsCases <- courtobs %>%
  nrow(.)

#total number of cases scheduled by court in time period of court observations
obsCourt <- evictions %>%
  filter(appearance_date >= as.Date("2022-11-01") & appearance_date < as.Date("2023-04-20")) %>%
  group_by(court) %>%
  summarize(totEvic = n()) %>%
  filter(court %in% c("Court  1-1",
                     "Court  1-2",
                     "Court  3-1",
                     "Court  3-2",
                     "Court  4-1",
                     "Court  5-1")) %>%
  ungroup(.)

#total number of cases observed by volunteers in observation period by court
jpCasesObs <- courtobs %>%
  group_by(jpCourt) %>%
  summarize(obsCases = n()) %>%
  mutate(perObsCases = obsCases/totObsCases,
         court = ifelse(jpCourt == "1-1", "Court  1-1",
                     ifelse(jpCourt == "1-2", "Court  1-2",
                     ifelse(jpCourt == "3-1", "Court  3-1",
                     ifelse(jpCourt == "3-2", "Court  3-2",
                     ifelse(jpCourt == "4-1", "Court  4-1",
                     ifelse(jpCourt == "5-1", "Court  5-1", "Other"))))))) %>%
  ungroup(.) %>%
  left_join(., obsCourt) %>%
  mutate(perAllCases = obsCases/totEvic)

#cases observed compared to entire county
jpCasesCty <- jpCasesObs %>%
  filter(jpCourt %in% c("1-1", "3-1", "5-1")) %>%
  summarize(obsCases = sum(obsCases),
            totEvic = sum(totEvic)) %>%
  mutate(perAllCases = round(obsCases/totEvic, digits = 3))

#case outcome by party filed in favor of by jp court
caseOutcome <- courtobs %>%
  filter(!is.na(jpCourt)) %>%
  filter(jpCourt != "5-2") %>%
  mutate(outcomeParty = ifelse(str_detect(jpRuled, "Plaintiff"), "Ruled in Favor of Landlord",
                                          ifelse(str_detect(jpRuled, "Defendant"), "Ruled in Favor of Tenant", 
                                                 ifelse(str_detect(jpRuled, "abated"), "Abated",
                                                        ifelse(str_detect(jpRuled, "dismissed"), "Dismissed", jpRuled))))) %>%
  group_by(jpCourt, outcomeParty) %>%
  summarize(count = n()) %>%
  mutate(perOutcome = count/sum(count)) %>%
  ungroup(.)

#case outcome by appearance in court
outcomeApp <- courtobs %>%
  filter(!is.na(jpCourt)) %>%
  filter(jpCourt != "5-2") %>%
  filter(defAppear != "Unknown") %>%
  mutate(outcomeParty = ifelse(str_detect(jpRuled, "Plaintiff"), "Ruled in Favor of Landlord",
                                          ifelse(str_detect(jpRuled, "Defendant"), "Ruled in Favor of Tenant", 
                                                 ifelse(str_detect(jpRuled, "abated"), "Abated",
                                                        ifelse(str_detect(jpRuled, "dismissed"), "Dismissed", jpRuled))))) %>%
  group_by(defAppear, outcomeParty) %>%
  summarize(count = n()) %>%
  mutate(perOutcome = count/sum(count)) %>%
  ungroup(.)

#case outcome by representation for tenant
outcomeRep <- courtobs %>%
  filter(!is.na(jpCourt)) %>%
  filter(jpCourt != "5-2") %>%
  filter(defRep == "Yes, an attorney was present for the defendant." | defRep == "No, the defendant did not have additional representation.") %>%
  mutate(outcomeParty = ifelse(str_detect(jpRuled, "Plaintiff"), "Ruled in Favor of Landlord",
                                          ifelse(str_detect(jpRuled, "Defendant"), "Ruled in Favor of Tenant", 
                                                 ifelse(str_detect(jpRuled, "abated"), "Abated",
                                                        ifelse(str_detect(jpRuled, "dismissed"), "Dismissed", jpRuled))))) %>%
  group_by(defRep, outcomeParty) %>%
  summarize(count = n()) %>%
  mutate(perOutcome = count/sum(count)) %>%
  ungroup(.)

#appearing for hearing
defAppear <- courtobs %>%
  filter(!is.na(jpCourt),
         defAppear != "Unknown") %>%
  group_by(jpCourt, defAppear) %>%
  summarize(count = n()) %>%
  filter(count > 1) %>%
  ungroup(.) %>%
  pivot_wider(names_from = defAppear, values_from = count, values_fill = 0) %>%
  mutate(AppearPer = Yes/(Yes+No),
         type = "Tenant (Defendant) Appeared for Hearing")

plaAppear <- courtobs %>%
  filter(!is.na(jpCourt),
         plaintiffDesignee != "Unknown") %>%
  group_by(jpCourt, plaintiffDesignee) %>%
  summarize(count = n()) %>%
  filter(count > 1) %>%
  ungroup(.) %>%
  pivot_wider(names_from = plaintiffDesignee, values_from = count, values_fill = 0) %>%
  mutate(AppearPer = Yes/(Yes+No),
         type = "Landlord (Plaintiff Appeared for Hearing")

caseAppear <- rbind(defAppear, plaAppear)

#case outcomes when defendant is present for all cases observed
defPresent <- courtobs %>%
  filter(!is.na(jpCourt),
         defAppear != "Unknown") %>%
  group_by(defAppear, jpRuled) %>%
  summarize(count = n()) %>%
  ungroup(.) %>%
  pivot_wider(names_from = defAppear, values_from = count, values_fill = 0) %>%
  mutate(NoPer = No/sum(No),
         YesPer = Yes/sum(Yes))

#case outcomes when defendant has representation for all cases observed
defRepresent <- courtobs %>%
  filter(!is.na(jpCourt),
         defAppear == "Yes") %>%
  mutate(legalRep = ifelse(defRep == "Yes, an attorney was present for the defendant.", "Attorney present for tenant (defendant).", "No attorney present for tenant.")) %>%
  group_by(jpCourt, legalRep) %>%
  summarize(count = n()) %>%
  mutate(perRep = count/sum(count)) %>%
  ungroup(.)

# notice to vacate 
ntv <- courtobs %>%
  filter(!is.na(noticeVacate)) %>%
  group_by(jpCourt, noticeVacate) %>%
  summarize(count = n()) %>%
  mutate(perntv = count/sum(count)) %>%
  ungroup(.)

# average hearing time
hearingtime <- courtobs %>%
  group_by(observationDate, observer) %>%
  arrange(timeStamp)

#eviction reason
evicReason <- courtobs %>%
  filter(!is.na(jpCourt)) %>%
  mutate(evicClean = ifelse(str_detect(evicReason, ","), "Multiple Reasons", evicReason)) %>%
  group_by(jpCourt, evicClean) %>%
  summarize(count = n()) %>%
  mutate(perCases = count/sum(count)) %>%
  ungroup(.)
```
# SUMMER 2023 FINDINGS

From `r startDate` through `r endDate`, `r smuStudents` SMU Law student volunteers observed
`r totObsCases` cases: `r jpCasesObs %>% filter(jpCourt == "1-1") %>% pull(obsCases)` in Justice of the Peace Court 1-1, `r jpCasesObs %>% filter(jpCourt == "3-1") %>% pull(obsCases)` in Justice of the Peace Court
3-1, and `r jpCasesObs %>% filter(jpCourt == "5-1") %>% pull(obsCases)` in Justice of the Peace Court 5-1. The `r totObsCases` observed cases represent `r paste0(jpCasesCty$perAllCases*100, "%") `
coverage of all initially scheduled hearings in 1-1, 3-1, and 5-1.

```{r, Cases Chart}
jpCasesObs %>%
  filter(!is.na(jpCourt) & obsCases > 1) %>%
  select(jpCourt, court, perAllCases) %>%
  mutate(perAntCases = 1-perAllCases) %>%
  pivot_longer(cols = c("perAllCases", "perAntCases")) %>%
  rename(perCases = value) %>%
  mutate(type = ifelse(name == "perAllCases", "Observed", "Not Observed")) %>%
  select(-name) %>%
  ggplot(aes(x = "", y = perCases, fill = type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_cpal() +
  facet_grid(~court) +
  scale_fill_cpal(theme = "factor") +
#  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent of All Cases Observed")
  
```

```{r, Cases Table}
jpCasesObs %>%
  filter(!is.na(jpCourt) & obsCases > 1) %>%
  select(court, obsCases, totEvic) %>%
  pivot_longer(cols = -court, names_to = "type") %>%
  pivot_wider(names_from = court, values_from = value) %>%
  mutate(type = ifelse(type == "obsCases", "CASES OBSERVED",
                       ifelse(type == "totEvic", "SCHEDULED HEARINGS*", "ERROR"))) #%>%
#  datatable(options = list(paging = FALSE,
#                           searching = FALSE,
#                           info = FALSE
#                           ),
#            rownames = FALSE) %>%
#  formatCurrency(c("Court  1-1", "Court  1-2", "Court  3-1", "Court  3-2", "Court  4-1", "Court  5-1"),
#                 interval = 3,
#                 mark = ",",
#                 currency = "",
#                 digits = 0)
```

> *The ‘scheduled hearings’ indicator was calculated based on the hearing date set at the time of the eviction filing. If a hearing date was subsequently rescheduled for a date later than August 5, then the case is still included here in the ‘scheduled hearings’ count. Some scheduled hearings may have been dismissed in advance of the docket, so ‘coverage’ here is likely an underrepresentation of cases observed relative to the number of cases heard during a docket.

## Across the `r totObsCases` observed cases, we found:

```{r, Cases by Reason}
evicReason %>%
  filter(jpCourt != "5-2") %>%
  ggplot(aes(x = "", fill=evicClean, y=perCases)) + 
  geom_bar(stat="identity") +
  coord_polar(theta = "y") +
  theme_cpal() +
  facet_grid(~jpCourt) +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent of All Cases Observed") +
  scale_fill_cpal(theme = "factor")
  
```

> *The reason for an eviction case might be unknown if a default judgment was awarded to a landlord and the facts of the case were not discussed. Reasons are rounded to the nearest whole number, so totals may not equal 100%.

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: In Dallas County, millions of dollars for rent relief have been available throughout the
pandemic. Eviction cases due to nonpayment of rent might have been remedied had the renter and landlord
been aware of and able to easily access rent relief programs.

```{r, Representation for Cases}
caseAppear %>%
  ggplot(aes(x = type, y = AppearPer, fill = type)) +
    geom_bar(stat = "identity") +
    facet_grid(~jpCourt) +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Landlords or their representatives almost always appeared for the hearing, but tenants
appeared less than half the time.") +
  scale_fill_cpal(theme = "factor") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

> *In fall 2022, volunteers observed in Justice of the Peace Court 1-1. The arrow here shows change from fall 2022 data collection. E.g., in summer 2023, tenants appeared 48% of the time in JP 1-1, but in fall 2022, tenants appeared 59% of the time.

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: When a tenant does not appear for an eviction hearing, a default judgment is usually awarded to the landlord, without consideration of the facts of the case. Tenants may not appear in court because they must go to work, lack transportation, are unclear about how eviction cases proceed, or have already determined to move out.

```{r, Legal Representation Tenants}
defRepresent %>%
  filter(jpCourt != "5-2") %>%
  ggplot(aes(x = legalRep, y = perRep, fill = legalRep)) +
    geom_bar(stat = "identity") +
    facet_grid(~jpCourt) +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Tenants rarely had legal representation.") +
  scale_fill_cpal(theme = "factor") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

> *In Spring 2023 in JP 1-1, tenants had legal representation XX% of the time.

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: : Unlike criminal cases, defendants in civil cases generally do not have a right to counsel
and are not assigned a public defender to represent them in court. As a result, very few tenants in eviction cases have legal representation and must defend themselves. Renters with legal representation in eviction hearings are much more likely to win their case and remain housed. Attorneys may also help renters negotiate terms with their landlords, keep eviction judgments off their records, and navigate other housing
supports, like rent relief.

```{r, notice to vacate}
ntv %>%
  filter(jpCourt != "5-2",
         noticeVacate == "Yes") %>%
  ggplot(aes(x = noticeVacate, y = perntv)) +
    geom_bar(stat = "identity",
             fill = "#008097") +
    facet_grid(~jpCourt) +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Tenants rarely had legal representation.") +
  scale_fill_cpal(theme = "factor") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

> *In Spring 2023 in JP 1-1, a Notice to Vacate was confirmed XX% of the time.

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: Unless a lease agreement states otherwise, Texas law requires a landlord to give a renter at least three days to move out, and the landlord must provide this notice (called a Notice to Vacate) in writing. An eviction cannot be filed in court before providing a Notice to Vacate. If a Notice to Vacate was not properly administered, then the renter may have recourse to have the case dismissed. If the Notice to Vacate is not confirmed during an eviction hearing, then there may be a missed opportunity to dismiss the eviction suit.

**TABLE OF AVERAGE HEARING TIMES**

> Average hearing length calculated by (1) counting the number of cases observed in each docket, including default judgments, (2) calculating the total time of the docket (end time - start time), (3) calculating the average length of each hearing by dividing docket time by number of cases observed, and then (4) calculating a weighted average across all observed dockets. Because not all cases in a single docket may have been
observed, average length of time here is likely conservative – i.e., average length of time per hearing is likely an overestimate.

```{r, eviction outcome}
caseOutcome %>%
  ggplot(aes(x = "", fill=outcomeParty, y=perOutcome)) + 
  geom_bar(stat="identity") +
  coord_polar(theta = "y") +
  theme_cpal() +
  facet_grid(~jpCourt) +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Across all observed cases, the judge ruled in favor of the landlord in XX% of cases and in
favor of the tenant in YY% of cases.") +
  scale_fill_cpal(theme = "factor")

```

> Outcomes are rounded to the nearest whole number, so totals may not equal 100%.

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: 

> Outcomes are rounded to the nearest whole number, so totals may not equal 100%

{{< fa house-chimney size=lg title="Housing" >}} **Why this matters**: Evicted low-income renters face a host of consequences: eviction increases the risk of homelessness, elevates long-term residential instability, and increases emergency room use. Eviction is also a matter of public record and can appear on public consumer reports (even if a renter won an eviction case), so evicted renters have a harder time finding a new place to live. Rent or court fees may also go to collections, negatively impacting a renter’s credit score.

## However, appearing in court and having legal representation matters for outcomes.When tenants appeared for their hearings, cases were decided in favor of the landlord 57% of the time.

```{r, outcomes with appearance}
outcomeApp %>%
  ggplot(aes(x = defAppear, fill=outcomeParty, y=perOutcome)) + 
  geom_bar(stat="identity", width = 0.5) +
  theme_cpal() +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Case outcomes when tenant/defendant appeared versus when tenant/defendant did not
appear (across all observations).") +
  scale_fill_cpal(theme = "factor") +
  coord_flip()

```

## When tenants had legal representation, cases were decided in favor of the landlord 10% of the time.

```{r, outcomes with representation}
outcomeRep %>%
  ggplot(aes(x = defRep, fill=outcomeParty, y=perOutcome)) + 
  geom_bar(stat="identity", width = 0.5) +
  theme_cpal() +
  theme_cpal() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Case outcomes when tenant/defendant had legal representation versus when
tenant/defendant did not have legal representation (across all observations)") +
  scale_fill_cpal(theme = "factor") +
  coord_flip()

```

# ABOUT CPAL
CPAL is a Dallas-based nonprofit organization committed to cutting child poverty in
Dallas, Texas in half within a generation by (1) generating relevant data insights to break big problems into small, actionable ones, (2) facilitating cross-sector, collective action based on data insights, and (3) applying principles of human-centered design to solve problems in partnership with communities. One of CPAL’s five areas of focus is housing, and CPAL’s long-term aspiration is to ensure all Dallas families have access to stable housing without worry of an involuntary move.
